Extreme rainfall in Paraguay during the 2015-2016 austral summer: causes and subseasonal-to-seasonal predictive skill

### 2017-09-18 03Writeup/paper.tex
Work on writeup

\documentclass[twocol]{ametsoc}
\journal{mwr}

@@

%Packages
\usepackage{
	siunitx, % units
	cleveref,% cross-references
	physics, % better physics notation
	url, % for URLs
	amsmath, % for math
}

% TODO Notes -- DO NOT KEEP FOR FINAL

@@

%\title{Causes and subseasonal-to-seasonal model skill of the persistent intense rainfall and flooding in Paraguay during the austral summer 2015-2016}
\title{Extreme rainfall in Paraguay during the 2015-2016 austral summer: causes and subseasonal-to-seasonal predictive skill}
\authors{James Doss-Gollin\correspondingauthor{James Doss-Gollin. Dept. of Earth and Environmental Engineering, Columbia University, 500 W. 120th St., New York, NY. USA.}}
\affiliation{Dept. of Earth and Environmental Engineering, Columbia University, 500 W. 120th St., New York, NY. USA \\ Columbia Water Center, Columbia University, 500 W. 120th St., New York, NY. USA}
\email{james.doss-gollin@columbia.edu}
\extraauthor{\'Angel G. Mu\~noz}
\extraaffil{Atmospheric and Oceanic Sciences (AOS). Princeton University. Princeton, NJ 08540-6649. USA  \\
	International Research Institute for Climate and Society (IRI). The Earth Institute. Columbia University. Palisades, NY. 10964-1000. USA}
    \extraauthor{Simon J. Mason}
\extraaffil{International Research Institute for Climate and Society (IRI). The Earth Institute. Columbia University. Palisades, NY. 10964-1000. USA}
\extraauthor{Max Past\'{e}n}
\extraaffil{Direcci\'{o}n de Meteorolog\'{i}a e Hidrolog\'{i}a. Paraguay \\
	Facultad Polit\'ecnica. Universidad Nacional de Asunci\'{o}n. Paraguay}

@@

% Why this is worthy of study -- needs some work!
During the austral summer of 2015-16, repeated intense rainfall events led to severe flooding on the Paraguay River, displacing approximately \num{170000} people \citep{BrackenridgeDFO} and generating tremendous property damage.
The Paraguay River (\cref{fig:study-area}) is the primary tributary of the Paran\'{a} River and an important contributor to the La Plata basin, and its flooding directly impacts the Paraguayan cities of Asunci\'{o}n, Concepci\'{o}n, and Pilar, and the Argentine cities of Formosa and Corrientes.
Intense rainfall and flooding along the Paraguay River also has important implications for hydropower generation along the Paran\'{a} River, for agriculture, and for regional water resource management.
Until now, the drivers of the November-February (NDJF) 2015-16 rainfall events and the extent to which they could have been skilfully predicted has not been analyzed in the scientific literature.
The goal of this paper is to address these issues.

@@

Probabilistic forecasts are normally used to convey information about the uncertainties associated to the prediction of interest.
It is a common practice to present these forecasts as probabilities for occurrence of below-normal, normal and above-normal precipitation, where the normal is defined as the long-term average of at least 20 years.
Here we use the period 1981-2010 for seasonal forecasts, and 1995-2014 for sub-seasonal ones.
Nonetheless, since we analyze extreme event predictions (90th percentile exceedances), rather than terciles.
Furthermore, we present all predictions in terms of odds relative to the climatological odds:
\begin{equation}
odds_{r} \equiv \frac{\frac{p}{1-p}}{\frac{p_c}{1-p_c}}
\end{equation}
where $p$ and $p_c$ represent the forecast probability and the climatological probability, respectively.
This choice helps to easily visualize in the forecast maps locations where the odds of occurring an extreme rainfall event are considerably higher than expected.
In contrast, using probabilities in these forecasts requires that the reader compares values against the climatological probability.
Our choice thus simplifies the comparison in a way that  also tends to be better understood by decision makers.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table*}[h]
%
\caption{Model Output Statistics (MOS) methods used to correct the ECMF sub-seasonal forecasts. Spatial domain for predictand is always the same (\SI{39}{\degree S}-\SI{17}{\degree S}; \SI{66}{\degree W}-\SI{49}{\degree W}). Two initializations are used: Nov 12th and 16th, 2015.} \label{t1}
\begin{center}
%
\scalebox{0.8}{
\begin{tabular}{ccccccccc}
%
\topline
%
Model & Region (Predictor) & Final predictor(s) selected \\
%
\midline
%
 \emph{Raw} & (\SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W}) & \begin{tabular}{@{}c@{}}Ensemble mean, computed using members from the  \\ two initializations. No correction performed. \end{tabular} \\
%
%
 \emph{XLR} & (\SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W}) & \begin{tabular}{@{}c@{}}Ensemble mean, computed using members from the  \\ two initializations. \end{tabular}  \\
%
 \emph{HLXR} & (\SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W}) & \begin{tabular}{@{}c@{}} Ensemble mean and spread, computed using  \\ members from the two initializations.\end{tabular}  \\
%
  \emph{PCR} & (\SIrange{60}{0}{\degree S}; \SIrange{80}{30}{\degree W}) & \begin{tabular}{@{}c@{}} Linear combination of model's EOFs, computed using \\ both initializations as independent predictors (10 EOFs). \end{tabular} \\
%
  \emph{CCA} & (\SIrange{60}{0}{\degree S}; \SIrange{80}{30}{\degree W}) & \begin{tabular}{@{}c@{}} Canonical modes computed using \\ both initializations as independent predictors. (10 predictor EOFs, \\ 4 predictand EOFs, 4 canonical modes)\end{tabular}  \\
%
\botline
%
\end{tabular}
}
%
\end{center}
%
\end{table*}
%
%

@@

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/StudyArea.pdf}
	\caption{
		The study area.
		(L): all of South America.
		The weather typing described in \cref{sec:methods,sec:weather-types} was calculated using geopotential height within the area indicated by the red box.
		The ``Lower Paraguay River Basin'', over which rainfall is averaged to create plots in \cref{fig:klee,fig:chiclet} and define intense rainfall days in \cref{fig:lagged-rain}, is shown by the black box.
		(R): The Paraguay River.
		As for (L), the ``Lower Paraguay River Basin'' over which rainfall is averaged is indicated.
		Streamflows shown in \cref{fig:streamflow} were taken from Asunci\'on, Paraguay, shown by the black dot.
		The Paraguay River and its tributaries, from the Natural Earth database (\texttt{www.naturalearthdata.com}), are also shown.
	}
  \label{fig:study-area}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in,height=9in,keepaspectratio=true]{../_figs/WTComposite.pdf}
	\caption{
		Circulation anomalies associated with each weather type.
    Weather types are separated by column.
		Top row gives anomalies of \SI{200}{\hecto\pascal} wind; arrows indicate the direction, and both shaded contours and arrow length indicate the magnitude of the wind anomaly.
    Second row shows circulation at \SI{850}{\hecto\pascal}; shaded contours give anomalies of geopotential height and arrows indicate the wind speed and direction.
    Bottom row shows monthly rainfall anomalies in units of \si{\milli\meter\per\day}.
	}
	\label{fig:wt-composite}
\end{figure*}

\begin{figure*}	\noindent\includegraphics[width=6.5in]{../_figs/LaggedRain.pdf}
	\caption{
		Circulation anomalies associated with intense rainfall (99th percentile exceedance of area-averaged rainfall) in the Lower Paraguay River region. Lagged composites are shown, by column, for $t = $ \SIlist{-2;-1;0;1}{\day} relative to the date of intense rainfall.
		Top row gives anomalies of \SI{200}{\hecto\pascal} wind; arrows indicate the direction, and both shaded contours and arrow length indicate the magnitude of the wind anomaly.
    Second row shows circulation at \SI{850}{\hecto\pascal}; shaded contours give anomalies of geopotential height and arrows indicate the wind speed and direction.
    Bottom row shows rainfall anomalies in units of \si{\milli\meter\per\day}.
	}
  \label{fig:lagged-rain}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/RainfallWeatherType}
	\caption{
		Weather type occurrence during two years.
		Time series of area-averaged rainfall (shown by the box in \cref{fig:study-area}) and weather type for each day of NDJF  2015-16.
		Dashed lines indicate the climatological 50th, 90th, and 99th percentiles of NDJF area-averaged rain.
	}
  \label{fig:klee}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/NDJF201516Anomaly.pdf}
	\caption{
    	Monthly circulation anomalies during NDJF 2015-16.
      Top row gives anomalies of \SI{200}{\hecto\pascal} wind; arrows indicate the direction, and both shaded contours and arrow length indicate the magnitude of the wind anomaly.
      Second row shows circulation at \SI{850}{\hecto\pascal}; shaded contours give anomalies of geopotential height and arrows indicate the wind speed and direction.
      Bottom row shows monthly rainfall anomalies, in unites of \si{\milli\meter\per\day}.
	}
  \label{fig:anomalies}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.4in]{../_figs/ForecastSkill.pdf}
	\caption{
		caption
	}
  \label{fig:subsprobfcst}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/Chiclet.pdf}
	\caption{
    Assessment of S2S model forecasts.
    Forecast time series from the ECMWF S2S model are shown for several lead times (\SIlist{2;10;25;40}{\day}) of each individual ensemble member (gray), the ensemble mean (black, dashed), and the observed (blue) rainfall.
		The observed rainfall is smoothed with a \SI{3.5}{\day} running mean in order to facilitate comparison between daily observations and the models, which are initialized every 3-4 days.
	}
  \label{fig:chiclet}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{Streamflow.pdf}
	\caption{
		River stage (height; in \si{\meter}) for the Paraguay River at four gauges along the Paraguay River.
		(a) Seasonality (black) and time series of 2015-16 observations (orange) at each stream gauge.
			The seasonality was fit using local polynomial regression as implemented in the \texttt{locfit} package in
			the \textbf{R} statistical programming environment \citep{Loader1999}.
		(b) Time series of daily stage measurements from 1929 to 2016.
	}
  \label{fig:streamflow}
\end{figure*}
\end{document}

>>>>

\documentclass{ametsoc}
\journal{mwr}

@@

%Packages
\usepackage{
	siunitx, % units
	cleveref,% cross-references
	physics, % better physics notation
	url, % for URLs
	amsmath, % for math
	array, % for fixed-width tables
}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\title{Extreme rainfall in Paraguay during the 2015-2016 austral summer: causes and subseasonal-to-seasonal predictive skill}
\authors{James Doss-Gollin\correspondingauthor{James Doss-Gollin. Dept. of Earth and Environmental Engineering, Columbia University, 500 W. 120th St., New York, NY. USA.}}
\affiliation{Dept. of Earth and Environmental Engineering, Columbia University, 500 W. 120th St., New York, NY. USA \\ Columbia Water Center, Columbia University, 500 W. 120th St., New York, NY. USA}
\email{james.doss-gollin@columbia.edu}
\extraauthor{\'Angel G. Mu\~noz}
\extraaffil{Atmospheric and Oceanic Sciences (AOS). Princeton University. Princeton, NJ 08540-6649. USA  \\
	International Research Institute for Climate and Society (IRI). The Earth Institute. Columbia University. Palisades, NY. 10964-1000. USA}
    \extraauthor{Simon J. Mason}
\extraaffil{International Research Institute for Climate and Society (IRI). The Earth Institute. Columbia University. Palisades, NY. 10964-1000. USA}
\extraauthor{Max Past\'{e}n}
\extraaffil{Direcci\'{o}n de Meteorolog\'{i}a e Hidrolog\'{i}a. Paraguay \\
	Facultad Polit\'ecnica. Universidad Nacional de Asunci\'{o}n. Paraguay}

@@

% THIS PARAGRAPH NEEDS SOME WORK
% COME BACK TO IT LATER
During the austral summer of 2015-16, repeated intense rainfall events led to severe flooding on the Paraguay River, displacing approximately \num{170000} people \citep{BrackenridgeDFO} and generating tremendous property damage.
The Paraguay River (\cref{fig:study-area}) is the primary tributary of the Paran\'{a} River and an important contributor to the La Plata basin, and its flooding directly impacts the Paraguayan cities of Asunci\'{o}n, Concepci\'{o}n, and Pilar, and the Argentine cities of Formosa and Corrientes.
Intense rainfall and flooding along the Paraguay River also has important implications for hydropower generation along the Paran\'{a} River, for agriculture, and for regional water resource management.
Until now, the drivers of the November-February (NDJF) 2015-16 rainfall events and the extent to which they could have been skilfully predicted has not been analyzed in the scientific literature.
The goal of this paper is to address these issues.

@@

Probabilistic forecasts are normally used to convey information about the uncertainties associated to the prediction of interest.
It is a common practice to present these forecasts as probabilities for occurrence of below-normal, normal and above-normal precipitation, where the normal is defined as the long-term average of at least 20 years.
Here we use the period 1981-2010 for seasonal forecasts, and 1995-2014 for sub-seasonal ones.
Nonetheless, since we analyze extreme event predictions (90th percentile exceedances), rather than terciles.
Furthermore, we present all predictions in terms of odds relative to the climatological odds:
\begin{equation} \label{eq:odds-ratio}
odds_{r} \equiv \frac{\frac{p}{1-p}}{\frac{p_c}{1-p_c}}
\end{equation}
where $p$ and $p_c$ represent the forecast probability and the climatological probability, respectively.
This choice helps to easily visualize in the forecast maps locations where the odds of occurring an extreme rainfall event are considerably higher than expected.
In contrast, using probabilities in these forecasts requires that the reader compares values against the climatological probability.
Our choice thus simplifies the comparison in a way that  also tends to be better understood by decision makers.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table*}[h]
%
\caption{
	Model Output Statistics (MOS) methods used to correct the ECMF sub-seasonal forecasts.
	Spatial domain for predictand is always the same (\SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W}).
	Two initializations are used: Nov 12th and 16th, 2015.} \label{tab:mos-methods}
\begin{center}
\begin{tabular}{L{0.75in} L{1.15in} L{3.5in}}
\hline\hline
Model & Region (Predictor) & Final predictor(s) selected \\
%
\hline
%
\emph{Raw} & \SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W} & Ensemble mean, computed using members from the  two initializations. No correction performed. \\
%
\emph{XLR} & \SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W} & Ensemble mean, computed using members from the  two initializations.  \\
%
\emph{HLXR} & \SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W} & Ensemble mean and spread, computed using  members from the two initializations.\\
%
\emph{PCR} & \SIrange{60}{0}{\degree S}; \SIrange{80}{30}{\degree W} & Linear combination of model's EOFs, computed using both initializations as independent predictors (10 EOFs).\\
%
\emph{CCA} & \SIrange{60}{0}{\degree S}; \SIrange{80}{30}{\degree W} & Canonical modes computed using both initializations as independent predictors. (10 predictor EOFs, 4 predictand EOFs, 4 canonical modes) \\
%
\hline\hline
%
\end{tabular}
\end{center}
\end{table*}

@@

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/StudyArea.pdf}
	\caption{
		Map of the study area.
		(L): all of South America.
		The domains of the ``Lower Paraguay River Basin'' and the domain used for weather typing are indicated in black and red, respectively.
		(R): The Paraguay River.
		As for (L), the ``Lower Paraguay River Basin'' is indicated.
		Streamflows shown in \cref{fig:streamflow} were taken from the fours stations indicated in green.
		The Paraguay River and its tributaries, from the Natural Earth database (\texttt{www.naturalearthdata.com}), are also shown.
	}
  \label{fig:study-area}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{./Streamflow.pdf}
	\caption{
		River stage (height; in \si{\meter}) for the Paraguay River at four gauges along the Paraguay River.
		(a) Seasonality (orange) and time series of 2015-16 observations (black) at each stream gauge.
			The seasonality was fit using local polynomial regression as implemented in the \texttt{locfit} package in
			the \textbf{R} statistical programming environment \citep{Loader1999}.
		(b) Time series of daily stage measurements from 1929 to 2016 at each station.
	}
  \label{fig:streamflow}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/NDJF201516Anomaly.pdf}
	\caption{
    	Monthly composite anomalies observed during NDJF 2015-16.
      Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
      Bottom row shows monthly rainfall anomalies, in units of \si{\milli\meter\per\day}.
	}
  \label{fig:anomalies}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/PSI_EOF_Loadings.pdf}
	\caption{
    	Leading four EOF loadings of daily NDJF \SI{850}{\hecto\pascal} streamfunction $\psi$ over chosen domain.
	}
  \label{fig:eof-loading}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in,height=9in,keepaspectratio=true]{../_figs/WTComposite.pdf}
	\caption{
		Composite anomalies associated with each weather type (column).
    Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
    Bottom row shows precipitation anomalies in units of \si{\milli\meter\per\day}.
	}
	\label{fig:wt-composite}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/RainfallWeatherType}
	\caption{
		Time series of area-averaged rainfall in the Lower Paraguay River Basin (\cref{fig:study-area}) for each day of NDJF  2015-16 (lines).
		The weather types corresponding to each day are indicated in both color and adjacent label.
		Dashed lines indicate the climatological 50th, 90th, and 99th percentiles of NDJF area-averaged rain over the Lower Paragauy River basin.
	}
  \label{fig:rain-wt}
\end{figure*}

\begin{figure*}	\noindent\includegraphics[width=6.5in]{../_figs/LaggedRain.pdf}
	\caption{
		Circulation anomalies associated with intense rainfall (99th percentile exceedance of area-averaged rainfall in the Lower Paraguay River Basin shown in \cref{fig:study-area}).
		Lagged composites are shown, by column, for $t = $ \SIlist{-2;-1;0;1}{\day} relative to the date of intense rainfall.
		Top row shows anomalies of streamfunction $\psi$ at \SI{850}{\hecto\pascal}.
    Bottom row shows precipitation anomalies in units of \si{\milli\meter\per\day}.
	}
  \label{fig:lagged-rain}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.5in]{../_figs/Chiclet.pdf}
	\caption{
    S2S model forecasts of area-averaged rainfall over the Lower Paraguay River Basin (\cref{fig:study-area}).
    Forecast time series from the ECMWF S2S model are shown for several lead times (\SIlist{5;14;19;30}{\day}).
		Each gray dot represents the forecast of a single ensemble member issued for the target date ($x$-axis) of area-averaged rainfall, in \si{\milli\meter\per\day} ($y$-axis).
		The blue line connects the ensemble means, and the black line shows the observed rainfall.
		Because the model is initialized twice-weekly, for a given lead time forecasts will only exist for two in seven possible target dates;  consequently the rainfall is smoothed with a 3-day running mean to facilitate visual comparison.
	}
  \label{fig:chiclet}
\end{figure*}

\begin{figure*}
	\noindent\includegraphics[width=6.4in]{../_figs/ForecastSkill.pdf}
	\caption{
		Model forecasts and skill scores for the methods indicated in \cref{tab:mos-methods}.
		Top row shows the intense rainfall forecast for 1-7 December 2015 as the odds ratio of \cref{eq:odds-ratio} over the target domain.
		A value greater than 1 indicates that the model forecast greater-than-average odds of rainfall exceeding the 90th percentile.
		The second shows the 2AFC skill score for each grid cell; a value greater than 50 indicates that the model outperforms climatology.
	}
  \label{fig:subsprobfcst}
\end{figure*}

\end{document}
